Something about cetacean heart rates and ballistocardiograms

### 2021-09-05 analysis/paper/paper.Rmd
Add bradycardia analysis and figure for classifying jerk peaks

  pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
    # geom_vline(aes(xintercept = dt),
    #            data = filter(bcg_surge_lsr, bcg_beat), 
    #            color = "blue", alpha = 0.5) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
    labs(y = bquote(Surge~(m~s^{-2}))) +
    t +
    theme(legend.position = "none")

  # B: BPM

  pB <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
    geom_line() +
    geom_point(color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "Heart rate (bpm)") +
    expand_limits(x = range(bcg_sample$dt), y = 0) +
    t

  # C: Sway, heave

  bcg_sway_heave <- bcg_sample %>% 
    rename(Sway = sway_filt, Heave = heave_filt) %>% 
    pivot_longer(c(Sway, Heave), 
                 names_to = "axis",
                 values_to = "acceleration")

  pC <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
    # geom_vline(aes(xintercept = dt),
    #            filter(bcg_sway_heave, bcg_beat), 
    #            color = "blue", alpha = 0.5) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    facet_grid(axis ~ .) +
    labs(y = bquote(Acc~(m~s^{-2}))) +
    t

  # D: Jerk

  pD <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = bquote(Jerk~(m~s^{-3}))) +
    expand_limits(y = 0) +
    t

@@

>>>>

  pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
    labs(y = bquote(Surge~(m~s^{-2}))) +
    t +
    theme(legend.position = "none")

  # B: BPM
  pB <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
    geom_line() +
    geom_point(color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "Heart rate (bpm)") +
    expand_limits(x = range(bcg_sample$dt), y = 0) +
    t

  # C: Sway, heave
  bcg_sway_heave <- bcg_sample %>% 
    rename(Sway = sway_filt, Heave = heave_filt) %>% 
    pivot_longer(c(Sway, Heave), 
                 names_to = "axis",
                 values_to = "acceleration")

  pC <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    facet_grid(axis ~ .) +
    labs(y = bquote(Acc~(m~s^{-2}))) +
    t

  # D: Jerk
  pD <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = bquote(Jerk~(m~s^{-3}))) +
    expand_limits(y = 0) +
    t

@@

```{r bradycardia-relax}
bw10hz <- bw180905_53_10hz %>%
  mutate(
    dive_id = split_dives(dt, depth,
                          surface = 2, min_depth = 10, min_dur = 5 * 60),
    dive_norm = normalize_dives(dive_id)
  )
bw_400hz$dive_norm <- approx(bw10hz$dt, bw10hz$dive_norm, bw_400hz$dt)$y

bpm_data <- bw_400hz %>%
  drop_na(bcg_bpm)

bpm_ts <- RobustLinearReg::theil_sen_regression(bcg_bpm ~ dive_norm, bpm_data)

bpm_pred <- tibble(dive_norm = seq(0, 1, length.out = 100)) %>%
  mutate(bcg_bpm = predict(bpm_ts, newdata = .))

ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
  geom_point() +
  geom_line(data = bpm_pred, color = "blue", size = 1.5) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Dive time (normalized)",
       y = "Heartrate (bpm)") +
  expand_limits(y = 0) +
  theme_classic()
```

The magnitude of bradycardia was strongest at the start of dives and relaxed towards the end. On average, heartrate increased from `round(predict(bpm_ts, data.frame(dive_norm = 0)), 1)` bpm at the start of dives to `round(predict(bpm_ts, data.frame(dive_norm = 1)), 1)` bpm by the end (Theil-Sen regression, $p < 10^{-10}$).

```{r jerk-peaks, fig.cap="Major peak selection procedure for identifying heartbeats. A Even after smoothing, the norm of the jerk vector has peaks that do not correspond to heartbeats (e.g. 2, 4). B In height-prominence space, there are two clear clusters corresponding to valid heartbeats (upper right) and artifacts (lower left). C "}
# Visualize peak thresholding algorithm
jerk <- bw_400hz$njerk_smooth
jerk[is.na(bw_400hz$region_id)] <- 0
window_s <- 1.8
fs_hz <- 400

# Find peaks
peaks <- pracma::findpeaks(jerk, minpeakdistance = window * fs_hz)
jerk_peaks <- slice(bw_400hz, peaks[, 2]) %>% 
  mutate(
    height = peaks[, 1],
    prominence = peak_prominences(jerk, peaks[, 2]),
    distance = sqrt((max(height) - height)^2 + (max(prominence) - prominence)^2)
  ) %>% 
  arrange(dt)

# Using region 14 as an example
jerk_example <- filter(bw_400hz, region_id == 14)
peak_example <- jerk_peaks %>% 
  filter(between(dt, first(jerk_example$dt), last(jerk_example$dt))) %>% 
  mutate(i = row_number())

# Labeled peaks in jerk
pA <- ggplot(jerk_example, aes(dt, njerk_smooth)) +
  geom_line() +
  geom_point(data = peak_example) +
  ggrepel::geom_text_repel(aes(label = i), peak_example) +
  expand_limits(y = 0) +
  labs(y = bquote(Jerk~(m~s^{-3}))) +
  theme_classic()
# Peak height, prominence
pB <- ggplot(peak_example, aes(height, prominence)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = i), max.overlaps = Inf) +
  coord_fixed() +
  theme_classic()

# All heights, prominences (with threshold)

height_prom <- jerk_peaks %>% select(height, prominence)
peak_pca <- prcomp(height_prom, scale = FALSE)
pc1 <- peak_pca$x[, "PC1"]
pc1_density <- density(pc1)
pc1_peaks <- pracma::findpeaks(pc1_density$y)
pc1_valley <- pracma::findpeaks(-pc1_density$y[pc1_peaks[1, 2]:pc1_peaks[2, 2]])
pc1_thr <- pc1_density$x[pc1_valley[1, 2] + pc1_peaks[1, 2] - 1]
pc1_axis <- c(peak_pca$sdev[1], 0) %*% t(peak_pca$rotation) %>% 
  scale(center = -colMeans(height_prom), scale = FALSE)
pc2_axis <- c(0, peak_pca$sdev[2]) %*% t(peak_pca$rotation) %>% 
  scale(center = -colMeans(height_prom), scale = FALSE)
pc1_density_rotate <- cbind(pc1_density$x, pc1_density$y) %*% 
  t(peak_pca$rotation) %>% 
  scale(center = -colMeans(height_prom), scale = FALSE)
pc1_thr_rot <- c(pc1_thr, 0) %*% t(peak_pca$rotation) %>% 
  scale(center = -colMeans(height_prom), scale = FALSE)
slope <- (pc2_axis - peak_pca$center)[2] / (pc2_axis - peak_pca$center)[1]
intercept <- pc1_thr_rot[2] - slope * pc1_thr_rot[1]
pc1_angle <- atan2(peak_pca$rotation[2, 1], peak_pca$rotation[1, 1]) * 180 / pi
pc2_angle <- pc1_angle - 90

pC <- ggplot(jerk_peaks, aes(height, prominence)) +
  geom_point() +
  annotate("segment", 
           x = peak_pca$center[1], y = peak_pca$center[2], 
           xend = pc1_axis[1], yend = pc1_axis[2],
           color = "red", arrow = arrow(length = unit(0.30, "cm"))) +
  annotate("text", label = "PC1",
           x = pc1_axis[1], y = pc1_axis[2],
           hjust = -0.1, angle = pc1_angle,
           color = "red") + 
  annotate("segment", 
           x = peak_pca$center[1], y = peak_pca$center[2], 
           xend = pc2_axis[1], yend = pc2_axis[2],
           color = "red", arrow = arrow(length = unit(0.30, "cm"))) +
  annotate("text", label = "PC2",
           x = pc2_axis[1], y = pc2_axis[2],
           hjust = 1.1, angle = pc2_angle,
           color = "red") + 
  geom_abline(slope = slope, intercept = intercept, 
              color = "red", linetype = 2) +
  coord_fixed() +
  theme_classic()

# density along pc1
pD <- as_tibble(peak_pca$x) %>% 
  ggplot(aes(PC1)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(color = "blue") +
  geom_vline(xintercept = pc1_thr, color = "red", linetype = 2) +
  theme_classic()

cowplot::plot_grid(pA, pB, pC, pD,
                   nrow = 2,
                   labels = "AUTO")
```
